---
title: "Simulate Cell Type Confounding"
author: "Chris McKennan"
date: 2016-02-05
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```

The purpose of this is to simulate data from the model $Y_{p \times n} = B_{p \times d}X_{d \times n} + L_{p \times k}B_{k \times d}^*X + LW_{k \times n} + Z_{p \times k'}\Gamma_{k' \times n} + E$. We want to know if we can get a reasonable estimate for $B$ using our proposed methods. We will let $W$ be the matrix of cell types. To start off, we will assume the covariates $X$ are simply disease status indicators, i.e. $d = 2$. 

## Load required functions

```{r Load Packages}
library(minfi)
library('IlluminaHumanMethylation450kmanifest')
library('IlluminaHumanMethylation450kanno.ilmn12.hg19')
library('FlowSorted.Blood.450k')      ##Methylation data on 6 males, 10 cell types; RGset object
library('RefFreeEWAS')
library('nlme')
library('corpcor')
library('sva')
source("../R/SimulateData.R")
```


## Preliminary analysis to get an idea of effect sizes of $B$ and $L$.

Get the data into R and initialize global variables.

```{r Preliminary Input}
path.LPS.kids <- '/Users/Chris/Desktop/Uchicago/Nicolae/DNAMethylation/MichellesData/HTkids_cellprop/HTkids_LPS_covar_ImpAct_10115.txt'
path.methylation <- '/Users/Chris/Desktop/Uchicago/Nicolae/DNAMethylation/MichellesData/Hutterite_Methylation_data/Meth_covar_allsamp_impcells_10115.txt'
data.cov <- data.frame(read.table(path.methylation, sep="\t", quote="\"", dec=".", header=T, check.names=F))
data.LPS <- data.frame(read.table(path.LPS.kids, sep="\t", quote="\"", dec=".", header=T, check.names=F))
cell.names <- colnames(data.LPS)[(ncol(data.LPS) - 5):ncol(data.LPS)]
n.cells <- length(cell.names)
n.ind <- nrow(data.LPS)
ind.ids.cell <- data.LPS$Rowid    #IDs of individuals with cell type data
ind.ids.covar <- data.cov$Rowid   #IDs of individuals in complete covariate file
```

Load Michele's data

```{r Load Michelles Data}
#Load SWAN-normalized data
#Variables are Mset.swan.norm and rgset
load(file="/Users/Chris/Desktop/Uchicago/Nicolae/DNAMethylation/MichellesData/MyWork/SavedObjects/Rfile.RData")
meth.norm <- getMeth(Mset.swan.norm)
unmeth.norm <- getUnmeth(Mset.swan.norm)
beta.values <- meth.norm/(meth.norm+unmeth.norm+100)  #These are ordered by row in data.cov
```

Identify columns of the data that correspond to the data that I have in data.LPS (i.e. cell type data) and perform a regression. Note that I have not removed CHIP effect. I am only interested in getting an idea of the relative effects of cell type and asthma status.

```{r Identify Standardized Effects}
col.use <- match(ind.ids.cell, ind.ids.covar)
beta.cell <- beta.values[,col.use]
M.cell <- log(beta.cell/(1-beta.cell))
m <- nrow(M.cell); n <- ncol(M.cell)  #m = #CpG sites; n = #individuals
model.mat <- model.matrix(~Tcell.act + Bcell.act + Eos.act + Neut.act + Mono.act + asthma + sex, data=data.LPS)
beta.op <- solve(t(model.mat) %*% model.mat, t(model.mat))
H <- model.mat %*% beta.op
H.perp <- diag(1,nrow(H)) - H

coef.M <- beta.op %*% t(M.cell)
coef.beta <- beta.op %*% t(beta.cell)
var.mat <- solve(t(model.mat) %*% model.mat)
var.effects <- rep(NA, m)
stand.effects <- array(NA, dim=dim(coef.M))
for (p in 1:m) {
  tmp.site <- cbind(M.cell[p,])
  var.effects[p] <- t(tmp.site) %*% H.perp %*% tmp.site/(n - ncol(model.mat))
  stand.effects[,p] <- 1/sqrt(var.effects[p]) * diag(1/sqrt(diag(var.mat))) %*% cbind(coef.M[,p])
}
```

Below are plots of the standardized effects for asthma and the six cell types at each CpG site. What is important to notice is that the asthma effect is somewhat comparable to the cell type effects. That is, we are losing a lot of information if we assume that the effect size is negligable compared to cell type (i.e. if we assume that $BX \ll LB^* X$).

```{r Plot Histograms}
beta.disease <- stand.effects[7,]   #My disease effects. These are quite large in comparison to the cell effects
hist(beta.disease, xlab='Standardized Asthma Effect Coefficient at CpG site', main='M-values Regressed onto Sex, Asthma status and Cell type', breaks=90)
hist(stand.effects[8,], xlab='Standardized Sex Effect Coefficient at CpG site', main='M-values Regressed onto Sex, Asthma status and Cell type', breaks=90)
hist(stand.effects[2,], xlab='Standardized Tcell Coefficient at CpG site', main='M-values Regressed onto Sex, Asthma status and Cell type', breaks=90)
hist(stand.effects[3,], xlab='Standardized Bcell Coefficient at CpG site', main='M-values Regressed onto Sex, Asthma status and Cell type', breaks=90)
hist(stand.effects[4,], xlab='Standardized Eos cell Coefficient at CpG site', main='M-values Regressed onto Sex, Asthma status and Cell type', breaks=90)
hist(stand.effects[5,], xlab='Standardized Neutrophil Coefficient at CpG site', main='M-values Regressed onto Sex, Asthma status and Cell type', breaks=90)
hist(stand.effects[6,], xlab='Standardized Monocyte Coefficient at CpG site', main='M-values Regressed onto Sex, Asthma status and Cell type', breaks=90)
```

Note that the sex effect is very large. We would lose a lot of this information if we naively 
## Simulate CpG expression

To simulate data, I will use ONLY the covariate disease status and will simulate $B$ from a normal distribution with $\sigma^2$ approximately equal to $\text{Var}\left(\hat{B}\right)$ above. There will be 3 cell types, so the confounding dimension should be only 2. To start, I will use 1,000 CpG sites and 50 indivuduals. Note that $\frac{p}{n} \approx 400$ in real data, so we are giving the method generous numbers.

```{r Simulate Data}
n <- 50; p <- 1e3
p.disease <- 0.5
fract.B <- 1   ##B will be simulated from N(0, v), where v = sigma.celleffect/sqrt(pi)/fract.B. It controls the relative size of the methylation effect

x.asthma <- sort(rbinom(n, 1, p.disease))
X.sim <- rbind(rep(1, n), x.asthma)
coef.cell <- c(0.1, 0.1)   #asthma causes increases in both cell types (same for all genes)
p.cell <- c(0.35, 0.35)   #Without disease, this is the mean of both cell types
L <- 5*t(coef.M[c(2,4),sample((1:m), p, replace=T)])  ## p x k matrix
WX <- cbind(p.cell, coef.cell) %*% X.sim + rbind(0.08*rnorm(n), 0.08*rnorm(n))  ##W(X) ~ MN(B*X, I, I), a k x n matrix. This is the cell type for each individual, which is a function of the covariates (albeit a weak relationship)
CellEffect <- L %*% WX
sd.celleffect <- sd(apply(CellEffect, 1, mean))   #I want the asthma effect to be smaller than the average cell type effect
B <- sd.celleffect/sqrt(pi)/fract.B * rnorm(p)
M.main <- cbind(B) %*% rbind(x.asthma) + CellEffect
sd.all <- apply(M.main, 1, sd)
E <- sample(sd.all, p, replace=T)/2
M.sim <- M.main + diag(E) %*% matrix(rnorm(p*n), nrow=p, ncol=n)
```

## Try to estimate $B$ from the simulated data.
I will do this first by rotating by letting $X_{d \times n} = U \Sigma V^T$ and rotating M.sim by $V$ so that the last $n - d$ columns of M.sim do not depend on the covariates $X$, only the latent factors.

```{r Rotate}
svd.X <- svd(X.sim)
V.x <- svd.X$v    ##This is only n x d. We need to add to this
tmp.mat <- cbind(V.x, diag(n))
V.rotate <- qr.Q(qr(tmp.mat))   ##Complete V, where the first two columns are V.x
M.sim.rot <- M.sim %*% V.rotate   ##Rotate M.sim so that last n-d columns don't depend on X
d <- length(svd.X$d)
M.rot.L <- M.sim.rot[,(d+1):n]  ##This matrix should just be LWV + EV with nothing that depends on X
```

Use a simple SVD to estimate L and then project $Y$ onto $L_{\perp}$ to estimate $B$. Inependently permute the rows of M.rot.L to remove any hidden structure to estimate the latent dimension.
```{r Estimate B with naive svd}
eig.M <- svd(M.rot.L)
norm.eig.values.M <- eig.M$d^2/sum(eig.M$d^2)
p.values.eigs <- rep(0, n-d)
n.boot <- 100
for (b in 1:n.boot) {
  X.b <- PermMat(M.rot.L)
  eig.values.b <- svd(X.b)$d[1:(n-d)]
  norm.eig.values.b <- eig.values.b^2/sum(eig.values.b^2)
  ind.b <- as.numeric(norm.eig.values.M > norm.eig.values.b)

  p.values.eigs <- p.values.eigs + ind.b/n.boot
}
k <- which(p.values.eigs < 0.95)[1] - 1   ##Naive estimate of the latent dimension
naive.WV <- eig.M$v[,1:k]
naive.L <- eig.M$u[,1:k]
Relative.Fnorm <- norm(L %*% solve(t(L)%*%L, t(L)) - naive.L %*% solve(t(naive.L)%*%naive.L, t(naive.L)), "F")/ncol(L)   #Relative Frobenius norm

##Compute angle between subspaces##
tmp.eig <- svd(L %*% solve(t(L)%*%L, t(L)) - naive.L %*% solve(t(naive.L)%*%naive.L, t(naive.L)), nu=1, nv=1)
tmp.u <- tmp.eig$u[,1]
tmp.dist <- tmp.eig$d[1]
u.projL <- L %*% solve(t(L)%*%L, t(L)) %*% cbind(tmp.u); u.projL.naive <- naive.L %*% solve(t(naive.L)%*%naive.L, t(naive.L)) %*% cbind(tmp.u)
theta <- acos( (sum(u.projL*u.projL) + sum(u.projL.naive*u.projL.naive) - tmp.dist^2)/(2 * sqrt(sum(u.projL*u.projL)) * sqrt(sum(u.projL.naive*u.projL.naive))) )   ##Maximum angle between the two subspaces

##Project M.sim onto L.perp and estimate B##
Lnaive.perp <- diag(p) - naive.L %*% solve(t(naive.L)%*%naive.L, t(naive.L))
L.perp <- diag(p) - L %*% solve(t(L)%*%L, t(L))
M.sim.L.perp <- L.perp %*% M.sim
M.sim.Lnaive.perp <- Lnaive.perp %*% M.sim

B.operator <- solve(X.sim %*% t(X.sim), X.sim)
Bhat.L.perp <- M.sim.L.perp %*% t(B.operator)
Bhat.Lnaive.perp <- M.sim.Lnaive.perp %*% t(B.operator)
Bhat.noL <- M.sim %*% t(B.operator)
```

## Plot results for $\hat{B}$

```{r Plot B}
plot(Bhat.L.perp[,2], B, main='Estimating B Knowing Loading Matrix, L', xlab='OLS Estimate for B after Projecting Y onto L.perp', ylab="True B")
plot(Bhat.Lnaive.perp[,2], B, main='Estimating B by Estimating Loading Matrix, L.hat', xlab='OLS Estimate for B after Projecting Y onto L.hat.perp', ylab="True B")  ##Note that this is identical to what Houseman does...
plot(Bhat.noL[,2], B, main='Estimating B without Knowing Loading Matrix', xlab='OLS Estimate for B', ylab="True B")

```

## Session information

```{r info}
sessionInfo()
```
